---
title: "model-validation"
author: "diana baetscher"
date: "2025-06-03"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Based off of Charles's original analysis, which was using code from Nick Weber.
Here I've modified things a bit - updating the final model runs using the optimal alpha parameter value based on the analysis of alpha.
I'm also curious about the use of Pearson's correlations in addition to the glmnet weight coefficients for filtering informative loci.
I toyed with those a bit and basically assumed that we should keep all 12 loci (actually 11 because LOC114 had no usable CpG methylation) unless they are undermining the relationship between CpG methylation and age. I think this is slightly different than filtering that one might do with RADseq or WGS data for identifying a smaller number of age-informative CpG sites.




```{r load-libraries}

library(stringr)
library(paletteer)
library(ggplot2)
library(RcppParallel)
library(tidyr)
library(multidplyr)
library(emmeans)
library(coda)
library(igraph)
library(psych)
library(glmnetUtils)
library(glmnet)
library(Matrix)
library(Metrics)
```


```{r}
library(rstan)
library(rstanarm)
library(tidyverse)
```

```{r load-methylation-data}
dataM <- read_csv("outputs/SSL_singleplex_bismark_output.csv") # output from Bismark

# reformat into wide format, just the CpG methylation sites
dataM_wide <- dataM %>%
  mutate(beta = `Methylated CpGs`/(`Methylated CpGs`+`Unmethylated CpGs`)) %>%
  select(sample, locus, beta) %>%
  pivot_wider(names_from = locus, values_from = beta) %>%
  select(-LOC114) # these values are zeros and NAs

# read in and add metadata
meta <- read_csv("../data/Ejubatus_ABLG_export_20250506.csv")

# add ages to the methylation data
m_df <- dataM_wide %>%
  left_join(., meta, by = c("sample" = "ABLG")) %>%
  select(sample, AgeYRS, c(1:12))

# trimming first column on the dataframe
ids <- m_df$sample
methyl_values <- m_df[,-1]

```


```{r subset-training-and-test-data}
# 70/30 split

# Determine the number of rows and split indices
set.seed(42)
n_rows <- nrow(methyl_values)
train_indices <- sample(n_rows, size = (n_rows*0.7), replace = F)

# split the data
train_data <- methyl_values[train_indices, ]
test_data <- methyl_values[-train_indices, ]

# trimming first column on the testing and training dataset
# this removes the age information
# train_data <- train_data[,-1]
# test_data <- test_data[,-1]

```

### Modeling

Using the glmnet package, typical to do cross-validation on the training set to determine the optimal penalty parameter (lambda)
10-fold



First, work with the training dataset

Documentation here: https://glmnet.stanford.edu/reference/glmnet.html

```{r running-the-epigenetic-clock-model}
# this uses a cross-validated GLM
CVGLM <- glmnet::cv.glmnet(x = as.matrix(train_data[,-1]), # input matrix, minus response variable
                   y = as.matrix(train_data[,1]), # age information (response)
                   alpha = 0, # elastic net mixing parameter: can be empirically determined between 0-1
                   nfolds = nrow(methyl_values), # with this parameter, we're using leave-one-out cross-validation
                   type.measure = "mae",
                   family = "gaussian",
                   grouped = FALSE)

min(CVGLM$cvm)

CVGLM
```


```{r Loop-over-indivs-for-age-prediction}
z <- lapply(1:nrow(train_data), function(x){
  
  fit <- glmnet::glmnet(
                x = as.matrix(train_data[-x,-1]),
                y = as.matrix(train_data[-x,1]),
                alpha = 0.05,
                lambda = CVGLM$lambda.min, # lambda is derived from the training data model
                family="gaussian")

   pred <- predict(fit, as.matrix(train_data[x,-1]), type = "response")

  return(data.frame(pred, true = train_data[x,1]))
})

z <- do.call(rbind, z)

colnames(z) <- c("PredictedAge", "ChronologicalAge")
z 

```


```{r linear-regression-on-test-data}
regression_FCs <- lm(z$ChronologicalAge ~ z$PredictedAge)
summary(regression_FCs)

```

```{r}
#MAE (median absolute error) on testing dataset
mae(z$ChronologicalAge, z$PredictedAge)
```

```{r}
#Calculate residuals on testing dataset
z$Residuals <- z$PredictedAge - z$ChronologicalAge

#Calculate relative error on testing dataset
z$RelativeError <- abs(z$Residuals) / z$ChronologicalAge
regression_error <- lm(z$RelativeError ~ z$ChronologicalAge)
summary(regression_error)
```

```{r predictions}
# Predict
pred <- predict(CVGLM, newx = as.matrix(test_data[,-1]), s="lambda.min")

pred <- data.frame(pred)
colnames(pred) <- c("PredictedAge")
pred$ChronologicalAge <- test_data$AgeYRS

#Linear regression on test dataset
regression <- lm(pred$ChronologicalAge ~ pred$PredictedAge)
summary(regression)

regression$model
```

```{r mae-test-data}
#MAE of test dataset
test_mae <- round(mae(pred$ChronologicalAge, pred$PredictedAge), digits = 2)
```


```{r test-data-residuals}
#Calculate residuals on test dataset
pred$Residuals <- pred$PredictedAge - pred$ChronologicalAge

#Calculate relative error on test dataset
pred$RelativeError <- abs(pred$Residuals) / pred$ChronologicalAge

regression_error <- lm(pred$RelativeError ~ pred$ChronologicalAge)
summary(regression_error)
```

```{r diff-residuals-training-and-test}
# Test for difference in residuals between training and testing datasets
var.test(z$Residuals, pred$Residuals)
```

```{r}
ggplot(pred, aes(x = ChronologicalAge, y = PredictedAge)) +
  geom_point(color = "red") +  # Color points red
  geom_smooth(method = "lm", se = TRUE, color = "blue", fill = "lightblue") +  # Confidence envelope with line
  labs(x = "Chronological age", y = "Predicted age") +
  coord_fixed() +  # Ensures same scale for x and y
  annotate("text", x = 10, y = 22.5, label = "R-squared = 0.6678", # from the regression model
           color = "black", size = 3) + 
  annotate("text", x = 10, y = 21.5, label = paste0("Median absolute error = ", test_mae), 
           color = "black", size = 3) +
  theme_minimal()


```


Filter CpG sites using Pearson's correlations - Isn't this also what the elastic net is supposed to do?

```{r}
### Perfoming a pearson’s correlations between know ages and methylation levels
#Calculate Pearson's correlations
corr_matrix <- corr.test(x = as.matrix(methyl_values[,-1]), 
                         y = methyl_values[, 1], 
                         use="pairwise", 
                         method="pearson", 
                         adjust="BH", 
                         alpha=0.05)


```

```{r convert-matrix-to-dataframe}
#Format dataframe
corr_results <- data.frame(corr_matrix$r, corr_matrix$p.adj)
corr_results <- tibble::rownames_to_column(corr_results, "Loc")
colnames(corr_results) <- c("Loc", "r", "padj")
corr_results$absr <- abs(corr_results$r)

# loci with the top 10% of Pearson's correlations
quantile(corr_results$absr, 0.90)
# does this filtering scheme make it relative to other loci? 
# when we would be fine keep as many loci as are providing information for the model...
```

Rather than filtering for the top 10% of Pearson's correlations, I'll remove the loci with very low Pearson's correlations (< 0.1).

```{r remove-loci-with-low-correlations}
loc_highr <- corr_results %>%
  filter(absr > 0.1)
loc_percent_filt <- methyl_values %>%
  select(loc_highr$Loc, AgeYRS)

```
Rerun cv.glmnet with filtered data.
Does it make a difference?

```{r filtered-loci-based-on-Pearson-correlations}
CVGLM <- cv.glmnet(x = as.matrix(loc_percent_filt[,-length(loc_percent_filt)]),
                   y = as.matrix(loc_percent_filt[,length(loc_percent_filt)]),
                   nfolds = nrow(loc_percent_filt),
                   alpha = 0, # revisit the appropriate alpha value
                   type.measure = "mae",
                   family = "gaussian",
                   grouped=FALSE)

min(CVGLM$cvm)
```
That is a smaller MAE than when using all loci.


Filtering loci using glmnet weight coefficients

```{r}
# Get CpG site model coefficients
coefList <- coef(CVGLM, s=CVGLM$lambda.min)
coefList <- data.frame(coefList@Dimnames[[1]][coefList@i+1],coefList@x)
names(coefList) <- c('loc','val')
coefList <- coefList[-1, ]
colnames(coefList) <- c("Loc", "Weight")
coefList$AbsWeight <- abs(coefList$Weight)

```
```{r remove-loci-based-on-glmnet-weights}
# Remove CpG sites in bottom 10% of contributions
quantile(coefList$AbsWeight, 0.10)

```

**WHY USE TWO METHODS: first Pearson's correlationa and then the glmnet. Presumably, one could just use the GLMnet filter?

```{r glmnet-filter}
loc_highweights <- coefList %>%
  filter(AbsWeight > 17)

sealion_perc_filt <- methyl_values %>%
  select(loc_highweights$Loc, AgeYRS)

```


```{r re-run-cvglmnet-with-filtered-loci}
CVGLM <- cv.glmnet(x = as.matrix(sealion_perc_filt[,-(length(sealion_perc_filt))]),
                   y = as.matrix(sealion_perc_filt[,length(sealion_perc_filt)]),
                   nfolds = nrow(sealion_perc_filt),
                   alpha = 0, # revisit this parameter for appropriate tuning
                   type.measure = "mae",
                   family = "gaussian",
                   grouped=FALSE)

min(CVGLM$cvm)
```
That MAE is smaller than the prior filter - but it's strange that the absolute weight is providing much info for PBX3


### alpha parameter

Identify the optimal alpha value
```{r choosing-alpha-parameter}
x <- as.matrix(sealion_perc_filt[,-length(sealion_perc_filt)])
y <- as.matrix(sealion_perc_filt[,length(sealion_perc_filt)])

sealion_glmnet <- list()
nreps <- 10
sealion_glmnet <- lapply(1:nreps, function(i){
fit <- cva.glmnet(x = x, 
                  y = y, 
                  nfolds=nrow(sealion_perc_filt), 
                  type.measure="mae", 
                  family="gaussian", 
                  grouped=FALSE)
    sealion_glmnet$model <- fit$modlist
})

# Rename lists
prefix <- "model"
suffix <- seq(1:10)
prefix2 <- "alpha"
suffix2 <- c(0.000, 0.001, 0.008, 0.027, 0.064, 0.125, 0.216, 0.343, 0.512, 0.729, 1.000)

for (i in 1:length(sealion_glmnet)) {
  names(sealion_glmnet) <- paste(prefix, suffix, sep="_")
    for (j in 1:11) {
  names(sealion_glmnet[[i]]) <- paste(prefix2, suffix2, sep="_") 
    }
}

# Pull parameters of interest
vec <- data.frame(matrix(ncol=2))
names(vec) <- c("min.lambda", "min.mae")

for (i in 1:length(sealion_glmnet)) {
  for (j in 1:length(sealion_glmnet[[i]])) {
    vec <- rbind(vec, c(sealion_glmnet[[i]][[j]]$lambda.min, min(sealion_glmnet[[i]][[j]]$cvm)))
    }
  }
vec <- vec[-1, ]
alpha <- c(0.000, 0.001, 0.008, 0.027, 0.064, 0.125, 0.216, 0.343, 0.512, 0.729, 1.000)
vec$alpha <- rep(alpha, 10)

```

```{r calculating-means-for-each-alpha}
#Calculate means for each alpha value
min.mae.means <- aggregate(vec$min.mae ~ as.factor(vec$alpha), FUN=mean)
min.lambda.means <- aggregate(vec$min.lambda ~ as.factor(vec$alpha), FUN=mean)
vec.means <- cbind(min.mae.means, min.lambda.means)
vec.means <- vec.means[ ,-3]
colnames(vec.means) <- c("alpha", "mean.min.mae", "mean.min.lambda")

#Extract best parameter values
best.params <- vec.means[which.min(vec.means$mean.min.mae), ]
best.params

```

```{r plot-alpha}
#Plot of alpha versus min.mae for each model iteration
ggplot(vec, aes(x=as.factor(alpha), y=min.mae)) +
  geom_point() +
  labs(x="alpha", y="min. mae") +
  theme_minimal()
```

Note: the MAE isn't changing by much here.

## Rerunning the model with the filtered data and optimal alpha

```{r}
# 70% training and 30% test data
# Determine the number of rows and split indices
set.seed(44)
n_rows <- nrow(sealion_perc_filt)
train_indices <- sample(n_rows, size = (n_rows*0.7), replace = F)

# Split the data
train_data <- sealion_perc_filt[train_indices, ]
test_data <- sealion_perc_filt[-train_indices, ]

### trimming first column on the testing stand training dataset
#train_data <- train_data[,-1]
#test_data <- test_data[,-1]

# Verify
print(dim(train_data))

print(dim(test_data))

```


```{r run-model-with-filtered-data}
CVGLM <- cv.glmnet(x = as.matrix(train_data[,-length(train_data)]),
                   y = as.matrix(train_data[,length(train_data)]),
                   nfolds = nrow(train_data),
                   alpha = 0.008, # using the optimal value here based on the above evaluation of alpha
                   type.measure = "mae",
                   family = "gaussian",
                   grouped = FALSE)

min(CVGLM$cvm)
```

Basically the MAE value predicted in the optimal alpha analysis.

Looping over individuals to get predicted age estimates from ‘CVGLM’ generated with filtered data
```{r model-with-filtered-data-and-optimal-alpha}
z <- lapply(1:nrow(train_data), function(x){
  fit <- glmnet(
                x = as.matrix(train_data[-x,-length(train_data)]),
                y = as.matrix(train_data[-x,length(train_data)]),
                alpha = 0.008,
                lambda = CVGLM$lambda.min,
                family="gaussian")

   pred <- predict(fit, as.matrix(train_data[x,-length(train_data)]), type = "response")

  return(data.frame(pred, true = train_data[x,length(train_data)]))
})

z <- do.call(rbind, z)
colnames(z) <- c("PredictedAge", "ChronologicalAge")
z

```

```{r linear-regression-on-testing-data}
#Linear regression on testing dataset
regression_FCs <- lm(z$ChronologicalAge ~ z$PredictedAge)
summary(regression_FCs)

```
```{r}
#MAE on testing dataset
opt_model_mae <- round(mae(z$ChronologicalAge, z$PredictedAge), digits = 2)

opt_model_mae
```

```{r}
#Calculate residuals on testing dataset
z$Residuals <- z$PredictedAge - z$ChronologicalAge

#Calculate relative error on testing dataset
z$RelativeError <- abs(z$Residuals) / z$ChronologicalAge
regression_error <- lm(z$RelativeError ~ z$ChronologicalAge)
summary(regression_error)
```

```{r plot-optimized-model-output}
ggplot(pred, aes(x = ChronologicalAge, y = PredictedAge)) +
  geom_point(color = "red") +  # Color points red
  geom_smooth(method = "lm", se = TRUE, color = "blue", fill = "lightblue") +  # Confidence envelope with line - seems like the SE could be based on the data?
  labs(x = "Chronological age", y = "Predicted age") +
  coord_fixed() +  # Ensures same scale for x and y
  annotate("text", x = 9, y = 22.5, label = "R^2 = 0.6611", # I can't figure out how to call this directly from the model
           color = "black", size = 3) + 
  annotate("text", x = 9, y = 21.5, label = paste0("MAE = ", opt_model_mae), 
           color = "black", size = 3) +
  theme_minimal() +
  scale_x_continuous(limits = c(5, 25)) +
  scale_y_continuous(limits = c(5, 25)) +
  geom_abline(slope = 1, color = "black", linetype = "dotted") +
  theme(
    axis.title.x = element_text(margin = margin(t = 10)),
    axis.title.y = element_text(margin = margin(r = 10))
  ) +
  labs(title = "Steller sea lion epigenetic clock:\nCVGLM test data, 8 loci, 29 indiv")
  
ggsave("outputs/ssl_modeled_output.png", width = 8, height = 6)
```

