---
title: "singleplex_loci"
author: "diana baetscher"
date: "2025-05-19"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Quick look at methylation patterns in targeted loci for Steller Sea Lions


Bismarck output from Charles for 12 loci amplified and barcoded individually and then sequenced with the MiSeq V3 paired-end, 2x75 bp sequencing.


```{r load-libraries-and-data}
library(tidyverse)

metadata <- read_csv("../data/Ejubatus_ABLG_export_20250506.csv") %>%
  mutate(ABLG = as.character(ABLG))

data <- read_csv("../data/locus002_bismark_summary_report.csv")

sample_df <- data %>%
  mutate(File = str_replace(File, "/home/cgumbi/methylation/sea_lion/bismark_output/locus002/alignment/", "")) %>%
  mutate(File = str_replace(File, "_S*_*_L001_R1_trimmed_bismark_bt2_pe.bam", "")) %>%
  separate(File, into = c("sample", "locus")) %>%
  filter(!locus %in% c("S0", "SSL"))
  
```

```{r read-in-addl-locus}
loc114 <- read_csv("../data/Loci114_bismark_summary.csv") 

loc114_df <- loc114 %>%
  mutate(File = str_replace(File, "./", "")) %>%
  separate(File, into = c("sample", "toss"), sep = "_S") %>%
  select(-toss) %>%
  mutate(locus = "LOC114") %>%
  select(sample, locus, everything())

# and then combine with the sample df
ssl_bismark_output <- loc114_df %>%
  bind_rows(sample_df)
```
```{r sequencing-depth-across-loci}
ssl_bismark_output %>%
  ggplot(aes(x = sample, y = `Total Reads`)) +
  geom_point() +
  facet_wrap(.~locus) +
  theme(
    axis.text.x = element_blank()
  ) 
```
Plenty of sequencing, even though it varies by locus.

```{r on-target-reads}
ssl_bismark_output %>%
  ggplot(aes(x = sample, y = `Aligned Reads`/`Total Reads`)) +
  geom_point() +
  facet_wrap(.~locus) +
  theme(
    axis.text.x = element_blank()
  ) +
  labs(title = "SSL: on-target reads by locus")
  
```


```{r looking-at-CpG-sites}

ssl_bismark_output %>%
  ggplot(aes(x = sample, y = `Methylated CpGs`/(`Methylated CpGs` + `Unmethylated CpGs`))) +
  geom_point() +
  facet_wrap(.~locus, scales = "free") +
  theme(
    axis.text.x = element_blank()
  ) +
  labs(title = "SSL: Methylation at CpG sites")


```





Is methylation level correlated with age? Add some metadata to explore...

```{r add-metadata}
ssl_bismark_output %>%
  left_join(., metadata, by = c("sample" = "ABLG")) %>%
  ggplot(aes(x = AgeYRS, y = `Methylated CpGs`/(`Methylated CpGs` + `Unmethylated CpGs`))) +
  geom_point() +
  facet_wrap(.~locus) +
  theme(
    axis.text.x = element_text(angle = 90)
  ) +
  labs(title = "SSL: methylation at CpG sites")

```
Very different background methylation levels at each locus, which sort of obscures patterns within each locus.


```{r plot-all-loci}
ssl_bismark_output %>%
  left_join(., metadata, by = c("sample" = "ABLG")) %>%
  ggplot(aes(x = AgeYRS, y = `Methylated CpGs`/(`Methylated CpGs` + `Unmethylated CpGs`))) +
  geom_point() +
  facet_wrap(.~locus, scales = "free") +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5),
    axis.title.x = element_text(margin = margin(t = 10))
  ) +
  labs(title = "SSL: methylation at CpGs sites",
        y = "Methylated CpGs / CpG sites")

ggsave("outputs/Methylated_CpGs_by_locus.png", width = 10, height = 8)
```
Are the outliers consistent across loci?
```{r check-outliers}
ssl_bismark_output %>%
  filter(locus == "LOC1139" & `Methylated CpGs`/(`Methylated CpGs` + `Unmethylated CpGs`) > 0.2) %>%
  left_join(., metadata, by = c("sample" = "ABLG")) %>%
  ggplot(aes(x = AgeYRS, y = `Methylated CpGs`/(`Methylated CpGs` + `Unmethylated CpGs`), color = sample)) +
  geom_point() +
  facet_wrap(.~locus, scales = "free") +
  theme(
    axis.text.x = element_text(angle = 90)
  ) +
  labs(title = "SSL: methylated CpGs / methylated CpG + unmethylated CpG")

```

```{r check-outliers-across-loci}
ssl_bismark_output %>%
  mutate(outlier = ifelse(sample == "37207", "37207", "no")) %>%
  left_join(., metadata, by = c("sample" = "ABLG")) %>%
  ggplot(aes(x = AgeYRS, y = `Methylated CpGs`/(`Methylated CpGs` + `Unmethylated CpGs`), color = outlier)) +
  geom_point() +
  facet_wrap(.~locus, scales = "free") +
  theme(
    axis.text.x = element_text(angle = 90)
  ) +
  labs(title = "SSL: methylation at CpG sites",
       y = "Methylated CpGs / CpG sites")
  
```



```{r double-check-if-loc114-has-no-methylated-CpGs}
ssl_bismark_output %>%
  filter(locus == "LOC114") %>%
  left_join(., metadata, by = c("sample" = "ABLG")) %>%
  ggplot(aes(x = AgeYRS, y = `Methylated CpGs`/(`Methylated CpGs` + `Unmethylated CpGs`))) +
  geom_point() +
  facet_wrap(.~locus, scales = "free") +
  theme(
    axis.text.x = element_text(angle = 90)
  )
```


Output combined bismark output df for Simon, but return to the analysis later.

```{r output-csv}
ssl_bismark_output %>%
  write_csv("outputs/SSL_singleplex_bismark_output.csv")

```


Thinking about class biases in the ascertainment samples:

Do we have sex information for these individuals?

```{r}
ssl_bismark_output %>%
  left_join(., metadata, by = c("sample" = "ABLG")) %>%
  ggplot(aes(x = AgeYRS, y = `Methylated CpGs`/(`Methylated CpGs` + `Unmethylated CpGs`), color = Sex)) +
  geom_point() +
  facet_wrap(.~locus, scales = "free") +
  theme(
    axis.text.x = element_text(angle = 90)
  )
```

```{r summary-stats}
# avg reads by locus/indiv
ssl_bismark_output %>%
  summarise(mean_reads = mean(`Total Reads`), .by = locus) %>%
  arrange(desc(mean_reads))

```

```{r linear-models}
linear_models <- ssl_bismark_output %>%
  mutate(CpG_methylation = `Methylated CpGs`/(`Methylated CpGs` + `Unmethylated CpGs`)) %>%
  left_join(., metadata, by = c("sample" = "ABLG")) %>%
  group_by(locus) %>%
    do(broom::tidy(lm(AgeYRS~CpG_methylation, .))) %>%
  mutate(significant = ifelse(p.value < 0.05, "yes", "no"))
  

ssl_bismark_output %>%
  left_join(., metadata, by = c("sample" = "ABLG")) %>%
  left_join(., linear_models, by = "locus") %>%
  ggplot(aes(x = AgeYRS, y = `Methylated CpGs`/(`Methylated CpGs` + `Unmethylated CpGs`), color = significant)) +
  geom_point() +
  facet_wrap(.~locus, scales = "free") +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 90)
  ) +
  labs(y = "Methylated CpGs / CpG sites",
       color = "linear model\np-value < 0.05")

```


