---
title: "singleplex_loci"
author: "diana baetscher"
date: "2025-05-19"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Quick look at methylation patterns in targeted loci for Steller Sea Lions


Bismarck output from Charles for 11/12 loci

```{r load-libraries-and-data}
library(tidyverse)

metadata <- read_csv("../data/Ejubatus_ABLG_export_20250506.csv") %>%
  mutate(ABLG = as.character(ABLG))

data <- read_csv("../data/locus002_bismark_summary_report.csv")

sample_df <- data %>%
  mutate(File = str_replace(File, "/home/cgumbi/methylation/sea_lion/bismark_output/locus002/alignment/", "")) %>%
  mutate(File = str_replace(File, "_S*_*_L001_R1_trimmed_bismark_bt2_pe.bam", "")) %>%
  separate(File, into = c("sample", "locus")) %>%
  filter(!locus %in% c("S0", "SSL"))
  
```

```{r}
sample_df %>%
  ggplot(aes(x = sample, y = `Aligned Reads`/`Total Reads`)) +
  geom_point() +
  facet_wrap(.~locus) +
  theme(
    axis.text.x = element_blank()
  ) +
  labs(title = "SSL: on-target reads by locus")
  
```


```{r}
sample_df %>%
  ggplot(aes(x = sample, y = `Methylated CpGs`/(`Methylated CpGs` + `Unmethylated CpGs`))) +
  geom_point() +
  facet_wrap(.~locus, scales = "free") +
  theme(
    axis.text.x = element_blank()
  ) +
  labs(title = "SSL: methylated CpGs / methylated CpG + unmethylated CpG")


```

It doesn't look like a ton of variation in methylation level across samples within a locus. Let's look at HSPA2 closer as an example.

```{r}
sample_df %>%
  filter(locus == "HSPA2") %>%
  ggplot(aes(x = sample, y = `Methylated CpGs`/(`Methylated CpGs` + `Unmethylated CpGs`))) +
  geom_point() +
  facet_wrap(.~locus) +
  theme(
    axis.text.x = element_text(angle = 90)
  ) +
  labs(title = "SSL, locus HSPA2: methylated CpGs / methylated CpG + unmethylated CpG")


```

Actually, that's a decent spread: <25% methylated CpGs to >50% CpGs.

Is it correlated with age though?

```{r add-metadata}
sample_df %>%
  filter(locus == "HSPA2") %>%
  left_join(., metadata, by = c("sample" = "ABLG")) %>%
  ggplot(aes(x = AgeYRS, y = `Methylated CpGs`/(`Methylated CpGs` + `Unmethylated CpGs`))) +
  geom_point() +
  facet_wrap(.~locus) +
  theme(
    axis.text.x = element_text(angle = 90)
  ) +
  labs(title = "SSL, locus HSPA2: methylated CpGs / methylated CpG + unmethylated CpG")

```
Cool! Looks like there's information in that locus.

Let's look at the rest?

```{r plot-all-loci}
sample_df %>%
  left_join(., metadata, by = c("sample" = "ABLG")) %>%
  ggplot(aes(x = AgeYRS, y = `Methylated CpGs`/(`Methylated CpGs` + `Unmethylated CpGs`))) +
  geom_point() +
  facet_wrap(.~locus, scales = "free") +
  theme(
    axis.text.x = element_text(angle = 90)
  ) +
  labs(title = "SSL: methylated CpGs / methylated CpG + unmethylated CpG")

ggsave("outputs/Methylated_CpGs_by_locus.png", width = 10, height = 8)
```

```{r}
sample_df %>%
  filter(locus == "LOC1139" & `Methylated CpGs`/(`Methylated CpGs` + `Unmethylated CpGs`) > 0.2) %>%
  left_join(., metadata, by = c("sample" = "ABLG")) %>%
  ggplot(aes(x = AgeYRS, y = `Methylated CpGs`/(`Methylated CpGs` + `Unmethylated CpGs`), color = sample)) +
  geom_point() +
  facet_wrap(.~locus, scales = "free") +
  theme(
    axis.text.x = element_text(angle = 90)
  ) +
  labs(title = "SSL: methylated CpGs / methylated CpG + unmethylated CpG")

```

```{r}
sample_df %>%
  mutate(outlier = ifelse(sample == "37207", "37207", "no")) %>%
  left_join(., metadata, by = c("sample" = "ABLG")) %>%
  ggplot(aes(x = AgeYRS, y = `Methylated CpGs`/(`Methylated CpGs` + `Unmethylated CpGs`), color = outlier)) +
  geom_point() +
  facet_wrap(.~locus, scales = "free") +
  theme(
    axis.text.x = element_text(angle = 90)
  ) +
  labs(title = "SSL: methylated CpGs / methylated CpG + unmethylated CpG")
  
```

